Data exploration through dot-driven development\footnote{This work was supported by Google Digital News Initiative.}

### 2017-01-13 pivot/paper.tex
Typos

This means that our use of type providers is more akin to meta-programming or code generation with
one important difference -- the schema provided by the pivot type provider is potentially infinite
(as there are always more operations that can be applied). The implementation relies on the fact 
that type providers are integrated into the type system and types can be provided lazilly. This is 
also a new aspect of our formalization in Section~\ref{sec:foo}.

@@

\begin{itemize}
\item The filtering operation is written using indexing $[\ldots]$ while all other operations are
  written using member invocation with (optionally named) parameters. In the first case, we write
  an expression $\ident{olympics}[\str{Games}]==\str{Rio (2016)}$ returning a vector of Booleans
  while in the other, we specify a column name using $\ident{by}=\str{Gold}$. In other languges,
  a parameter can also be a lambda function specifying a predicate or a transformation.

@@

LISP is perhaps the best example of a language that unifies many distinct constructs using a single
form. In LISP, everything is an s-experssion, that is, either a list or a symbol. In contrast, 
a typical data processing language uses a number of distinct constructs including indexers (for 
range selection and filtering), method calls (for transformations) and named parameters (for further 
configuration). Consider filtering and sorting:
%
\begin{equation*}
\begin{array}{l}
\ident{data}[\ident{data}[\str{Games}]==\str{Rio (2016)}] \qquad\ball{1}\\
\ident{data}.\ident{filter}(\kvd{fun}~\ident{row}\rightarrow\ident{row}.\ident{Games}=\str{Rio (2016)})\qquad\ball{2}\\
\ident{data}.\ident{sort\_values}(\ident{by}=\str{Gold}, \ident{ascending}=\kvd{False})\qquad\ball{3}\\
\end{array}
\end{equation*}

@@

\subparagraph{Members, type providers, discoverability.}
When sorting \ball{1} the member that specifies how sorting is done includes the name of the
column. This is possible because the pivot type provider tracks the column names (see 
Section~\ref{sec:columns}) and provides members based on the available columns suitable
for use as sort keys. When filtering \ball{2}, the member \qident{Rio (2016)} is provided based
on the values in the data source (we discuss this furter in Section~\ref{sec:pivot-filter}).

@@

Using descriptive member names is only possible when the names are discoverable. The above
code could be executed in a dynamically-typed languge that allows custom message-not-understood
handlers, but it would be impossible to get the name right when writing it. Our approach 
relies on discovering names through auto-completion as discussed in Section~\ref{sec:analysis-auto}.

@@

Source code editors for object-based languages with nominal type sytems often provide 
auto-completion for members of objects. This combination works extremely well in practice; the
member list is a complete list of what might follow after typing ``dot'' and it can be easily
obtained for an instance of known type. The fact that developers can often rely on just typing
``dot'' and choosing an appropriate member led to a semi-serious phrase dot-driven development,
that we (equally semi-seriously) adopt in this paper.

@@

In Figure~\ref{fig:pivot-members}, we list a subset of the provided members that were used in the
example in Section~\ref{sec:intro}. After choosing \qident{sort data}, the user is offered 
the possible sorting keys and directions. After choosing \qident{group data}, the user first 
selects the grouping key and then can choose one or more aggregations that can be applied on
other columns of the group. Thus an enitre data transformation (such as choosing top 8 athletes
by the number of gold medals) can be constructed using dot-driven development.

@@

\subparagraph{The ``then'' pattern.}
Nesting is an essential programming construct and it may be desirable to support it directly in 
the language, but the ``then'' pattern lets us express nesting without langauge support. In both 
of the cases above, the nested structure is specified by selecting one or more members and then 
ending the nested structure using the \ident{then} member.

@@

Other data transformations could be type checked in a similar way, but there are two drawbacks.
First, row types and typestate (although relatively straightforward) make the host language more
complex. Second, rules such as (\emph{drop-col}) make auto-completion more difficult, because the
editor needs to understand the rules and calculate what members may be invoked. This is a distinct
operation from type checking and type infernece (which operate on complete programs) that needs 
to be formalized and implemented.

@@

\begin{equation*}
\begin{array}{rclll}
  D & = & \multicolumn{3}{l}{ 
    \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} 
  }\\[0.5em]
  e & = & \Pi_{f_1, \ldots, f_n} (e) &&\textnormal{Projection -- select specified column names}\\
   & \lsep & \sigma_\varphi (e) &&\textnormal{Selection -- filter rows by given predicate}\\
   & \lsep & \tau_{f_1\mapsto\omega_1, \ldots, f_n\mapsto\omega_n}(e) &&\textnormal{Sorting -- sort by specified columns}\\
   & \lsep & \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n} (e) &&\textnormal{Grouping -- group by and calculate aggregates}
   \\[0.5em]
  \omega & = & \ident{desc}\lsep\ident{asc}  &&\textnormal{Sort order -- descending or ascending}\\
  \rho & = & \ident{count}  &&\textnormal{Count number of rows in the group}\\
   & \lsep & \ident{sum}~f  &&\textnormal{Sum numerical values of the column $f$}\\
   & \lsep & \ident{dist}~f &&\textnormal{Count number of distinct values of the column $f$}\\
   & \lsep & \ident{conc}~f &&\textnormal{Concatenat string values of the column $f$}\\
\\
\end{array} 
\end{equation*}

@@

The semantics of the algbra is given in Figure~\ref{fig:foo-relsem}. We use vector-based semantics
to support sorting and duplicate entries, but otherwise the formalization captures the usual
behaviour. In projection and sorting, we write $f_{p(1)}, \ldots, f_{p(m)}$ to refer to a selection
of fields from $f_1, \ldots, f_n$. Assuming $m\leq n$, $p$ can be seen as a mapping from $\{1\ldots m\}$
to a subset of $\{1\ldots n\}$. In selection, $\varphi$ is a predicate applied to a mapping from 
column names to values. In sorting, we assume that there is a permutation on row indices $q$ such
that the tuples obtained by selecting values according to the given sort key are ordered. The 
auxiliary definition $u_{k,l}$ negates the number to reverse the sort order when descending order
is required.

The most complex operation is grouping. We need to group data by the value of the column $f_i$ and
then apply aggregations $\rho_1,\ldots,\rho_m$. To do this, we first obtain a set of groups $g_1,\ldots, g_s$
where each group represents a set of indicies of rows belonging to each group. For a given group 
$g_i$ we can then obtain values of column $j$ for rows in the group as $\{v_{j,k}\,|\, k\in g_i\}$.
This is used to calculate the resulting data set -- the field $f_i$ becomes a new column formed
by the group keys (obtained by picking one of the indices from $g_j$ for each group); other fields
are calculated by aggregating data in various ways -- $|g_i|$ gives the number of rows in the group,
$\Sigma$ sums numerical values and $\Pi$ (a slight notation abuse) concatenates string values.

@@

\subparagraph{Reduction rules.} The reduction relation $\leadsto_L$ is prameterized by a function
$L$ that maps class names to class definitions, together with nested classes associated with the
class definition (used during type checking as discussed below). The map is not used in the 
reduction rules for the relational algebra, given in Figure~\ref{fig:foo-relsem} and so it was 
omitted there.

@@

A type provider is an executable component called by the compiler and the editor to 
provide information about types on demand. In our formalisation, we follow the 
style of Petricek et al. \cite{fsdata}, but we add laziness as discussed in Section~\ref{sec:foo-foo}.
We model the core operations (dropping columns, grouping and sorting) in Section~\ref{sec:pivot-core}
and refine the model to include filtering Section~\ref{sec:pivot-filter}.
For simplicity we omit paging, which does not affect the shape of data.

@@

A type provider is a function that takes static parameters, such as schema of the input data set, 
and returns a class name $C$ together with a mapping that defines the body of the class and 
definitions of delayed classes $L$ that may be used by the members of the class $C$. 
In our case, the schema $F$ is a mapping from field names to field types:
%
\begin{equation*}
\begin{array}{l}
\ident{pivot}(F) = C, \{ C \mapsto (\kvd{type}~C(x:\ident{Query}) = \ldots,L) \} \quad \textnormal{where}~F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}
\end{array}
\end{equation*}
%
The class $C$ provided by the pivot tpe provider has a constructor taking \ident{Query}, which 
represents the, possibly already partly transformed, input data set. It generates members that
allow the user to refine the query and access the data. The type provider is defined using several
helper functions discussed in the rest of this section.

@@

If we were using the relational algebra formalised in Section~\ref{sec:foo-rel} to construct 
queries, we can write an invalid program, e.g.~by attempting to select
a column $f$ using $\Pi_{f}$ from a data set that does not contain the column. This is not an issue 
when using the pivot type provider, because the provided types allow the user to construct only 
correct data transformations.

To formalise this, we prove partial soundness of the Foo calculus (Theorem~\ref{thm:foo-sound}),
which characterises the invalid programs that can be written using the \ident{Query}-typed 
expressions and then prove safety of the pivot type provider (Theorem~\ref{thm:pivot-safe}),
which shows that such errors do not occur when using the provided types.

\subparagraph{Foo calculus.} The Foo calculus consists of the relational algebra and simple object
calculus where objects can be constructed and their members accessed. It permits recursion as a
member can invoke itself on a new object instance. To accommodate this, we formalise soundness 
using progress (Lemma~\ref{thm:foo-progress}) and preservation (Lemma~\ref{thm:foo-pres}).

@@

The example given in Section~\ref{sec:intro} obtained top 8 athletes based on the number
of gold medals from Rio 2016. It used two operations that were omitted in the formalisation in
Section~\ref{sec:pivot-core}. We omitted paging to keep the host language simple, but we also
omitted filtering, which lets us write $\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}$.
This operation is worth further discussion. To support it, the type provider needs not only the
schema of the data set, but also sample data set that is used to offer the available values such 
as $\qident{Rio (2016)}$.

In the revised formalisation, the \ident{pivot} function which models the type provider takes the schema
$F$ together with sample data $D$ and provides the type with class context:
%
\begin{equation*}
\begin{array}{l}
\ident{pivot}(F, D) = C, L \quad \textnormal{where}\\
\qquad F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}\\
\qquad D=\{ f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} 
\end{array}
\end{equation*}

@@

\begin{figure}[b]
\begin{center}
\includegraphics[scale=0.30,trim=0mm 0mm 0mm 0mm,clip]{images/table.png} % left bottom right top
\end{center}
\caption{Athletes by the number of medals over the enitre history of Olympic games}
\label{fig:case-table}
\end{figure}

@@

We used The Gamma script with the pivot type provider to build an interactive web site 
(\url{rio2016.thegamma.net}) that visualizes a number of facts about Olympic medalists 
using the data set discussed in Appendix~\ref{app:olympics-csv} and used throughout this paper.
The web site lets the readers view and modify the source code and we also developed a number of
tools that make working with the source code easier, going beyond the basic auto-completion tooling
to enable dot-driven development as discussed in Section~\ref{sec:analysis-auto}. In this section,
we reivew our experience and outline some of the additional tools (available at 
\url{github.com/the-gamma}).

@@

The user interface lets the user choose aggregations to be caluclated over a group and select 
columns used for sorting. It is generated automatically by looking for specific pattern in the 
chain of member accesses -- we annotate members with annotations denoting whether a member is 
start of a list, list item or an end of a list. The editor then looks for parts of the chain of the
form $\qident{list start}.\qident{list item 1}.\qident{list item 2}.\qident{list end}$ and generates
a component that lets the user remove or add list items. An item cannot be removed if the operation
would break the code (e.g.~when it adds a member that is needed later) and items to be added
are chosen using available members (as in the standard auto-complete). The headers
shown in Figure~\ref{fig:case-opts} are provided as additional annotations attached to 
\qident{list start}.

@@

The Figure~\ref{fig:case-ed} shows the sample expression (discussed above) in the live 
editor\footnote{The live editor can be tested live as part of the documentation for the 
JavaScript package at \url{thegamma.net}}.
Note that the selected part of code is the \qident{by Gold descending} identifier and so the 
preview shows results as computed at that point of the query evaluation. Athletes with largest number
of gold medals appear first, but silver or bronze medals are not yed used as secondary sorting keys.
As the user moves through the code, or writes the code, the live preview is updated accordingly.

@@

\subparagraph{Human-computer intercation.}
We discussed how the pivot type provider simplifies the programming model (Section~\ref{sec:analysis}),
but it would be interesting to explore this aspect empirically through the perspective of HCI. 
The live editor shown in Section~\ref{sec:impl} offers a form of direct manipulation 
\cite{directman,directman2,dynamicq}. Unlike spreadsheets, we construct a transformation rather than
actually transforming data, which makes it more related to systems for query construction
\cite{spreadsheetalgebra,querydirect}. Our approach is somewhat different in that we see code 
as equally important to the direct manipulation interface.

@@

The pivot type provider offers a safe and easy to use layer over an underlying relational algebra 
that we use to model data transformations. As a key technical contribution of this paper, we 
formalise the type provider and prove that queries constructed using the types it provides are 
correct. Achieving this property by ohter means would require a language with complex
type system features such as typestate and row types.

@@

The data set used as an example in the case study discussed in Section~\ref{sec:impl} as well as
in the examples disussed throughout the paper is a single CSV file listing the entire history of
Olympic medals awarded since 1896. The data set can be found at \url{https://github.com/the-gamma/workyard}
together with scripts used to obtain it. The following is a representative example listing the
first 5 rows:
%
{\small
\begin{equation*}
\begin{array}{l}
\textbf{\sffamily {Games},{Year},{Discipline},{Athlete},{Team},{Gender},{Event},{Medal},{Gold},{Silver},{Bronze}}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Alfred Hajos},\kvd{HUN},\ident{Men},\ident{100m freestyle},\ident{Gold},\num{1},\num{0},\num{0}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Otto Herschmann},\kvd{AUT},\ident{Men},\ident{100m freestyle},\ident{Silver},\num{0},\num{1},\num{0}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Dimitrios Drivas},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors},\ident{Bronze},\num{0},\num{0},\num{1}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Ioannis Malokinis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors},\ident{Gold},\num{1},\num{0},\num{0}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Spiridon Chasapis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors},\ident{Silver},\num{0},\num{1},\num{0}\\
\end{array}
\end{equation*} }

>>>>

This means that our use of type providers is more akin to meta-programming or code generation with
one important difference -- the schema provided by the pivot type provider is potentially infinite
(as there are always more operations that can be applied). The implementation relies on the fact 
that type providers are integrated into the type system and types can be provided lazily. This is 
also a new aspect of our formalization in Section~\ref{sec:foo}.

@@

\begin{itemize}
\item The filtering operation is written using indexing $[\ldots]$ while all other operations are
  written using member invocation with (optionally named) parameters. In the first case, we write
  an expression $\ident{olympics}[\str{Games}]==\str{Rio (2016)}$ returning a vector of Booleans
  while in the other, we specify a column name using $\ident{by}=\str{Gold}$. In other languages,
  a parameter can also be a lambda function specifying a predicate or a transformation.

@@

LISP is perhaps the best example of a language that unifies many distinct constructs using a single
form. In LISP, everything is an s-expression, that is, either a list or a symbol. In contrast, 
a typical data processing language uses a number of distinct constructs including indexers (for 
range selection and filtering), method calls (for transformations) and named parameters (for further 
configuration). Consider filtering and sorting:
%
\begin{equation*}
\begin{array}{l}
\ident{data}[\ident{data}[\str{Games}]==\str{Rio (2016)}] \qquad\ball{1}\\
\ident{data}.\ident{filter}(\kvd{fun}~\ident{row}\rightarrow\ident{row}.\ident{Games}=\str{Rio (2016)})\qquad\ball{2}\\
\ident{data}.\ident{sort\_values}(\ident{by}=\str{Gold}, \ident{ascending}=\kvd{False})\qquad\ball{3}\\
\end{array}
\end{equation*}

@@

\subparagraph{Members, type providers, discoverability.}
When sorting \ball{1} the member that specifies how sorting is done includes the name of the
column. This is possible because the pivot type provider tracks the column names (see 
Section~\ref{sec:columns}) and provides members based on the available columns suitable
for use as sort keys. When filtering \ball{2}, the member \qident{Rio (2016)} is provided based
on the values in the data source (we discuss this further in Section~\ref{sec:pivot-filter}).

@@

Using descriptive member names is only possible when the names are discoverable. The above
code could be executed in a dynamically-typed language that allows custom message-not-understood
handlers, but it would be impossible to get the name right when writing it. Our approach 
relies on discovering names through auto-completion as discussed in Section~\ref{sec:analysis-auto}.

@@

Source code editors for object-based languages with nominal type systems often provide 
auto-completion for members of objects. This combination works extremely well in practice; the
member list is a complete list of what might follow after typing ``dot'' and it can be easily
obtained for an instance of known type. The fact that developers can often rely on just typing
``dot'' and choosing an appropriate member led to a semi-serious phrase dot-driven development,
that we (equally semi-seriously) adopt in this paper.

@@

In Figure~\ref{fig:pivot-members}, we list a subset of the provided members that were used in the
example in Section~\ref{sec:intro}. After choosing \qident{sort data}, the user is offered 
the possible sorting keys and directions. After choosing \qident{group data}, the user first 
selects the grouping key and then can choose one or more aggregations that can be applied on
other columns of the group. Thus an entire data transformation (such as choosing top 8 athletes
by the number of gold medals) can be constructed using dot-driven development.

@@

\subparagraph{The ``then'' pattern.}
Nesting is an essential programming construct and it may be desirable to support it directly in 
the language, but the ``then'' pattern lets us express nesting without language support. In both 
of the cases above, the nested structure is specified by selecting one or more members and then 
ending the nested structure using the \ident{then} member.

@@

Other data transformations could be type checked in a similar way, but there are two drawbacks.
First, row types and typestate (although relatively straightforward) make the host language more
complex. Second, rules such as (\emph{drop-col}) make auto-completion more difficult, because the
editor needs to understand the rules and calculate what members may be invoked. This is a distinct
operation from type checking and type inference (which operate on complete programs) that needs 
to be formalized and implemented.

@@

\begin{equation*}
\begin{array}{rclll}
  D & = & \multicolumn{3}{l}{ 
    \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} 
  }\\[0.5em]
  e & = & \Pi_{f_1, \ldots, f_n} (e) &&\textnormal{Projection -- select specified column names}\\
   & \lsep & \sigma_\varphi (e) &&\textnormal{Selection -- filter rows by given predicate}\\
   & \lsep & \tau_{f_1\mapsto\omega_1, \ldots, f_n\mapsto\omega_n}(e) &&\textnormal{Sorting -- sort by specified columns}\\
   & \lsep & \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n} (e) &&\textnormal{Grouping -- group by and calculate aggregates}
   \\[0.5em]
  \omega & = & \ident{desc}\lsep\ident{asc}  &&\textnormal{Sort order -- descending or ascending}\\
  \rho & = & \ident{count}  &&\textnormal{Count number of rows in the group}\\
   & \lsep & \ident{sum}~f  &&\textnormal{Sum numerical values of the column $f$}\\
   & \lsep & \ident{dist}~f &&\textnormal{Count number of distinct values of the column $f$}\\
   & \lsep & \ident{conc}~f &&\textnormal{Concatenate string values of the column $f$}\\
\\
\end{array} 
\end{equation*}

@@

The semantics of the algebra is given in Figure~\ref{fig:foo-relsem}. We use vector-based semantics
to support sorting and duplicate entries, but otherwise the formalization captures the usual
behaviour. In projection and sorting, we write $f_{p(1)}, \ldots, f_{p(m)}$ to refer to a selection
of fields from $f_1, \ldots, f_n$. Assuming $m\leq n$, $p$ can be seen as a mapping from $\{1\ldots m\}$
to a subset of $\{1\ldots n\}$. In selection, $\varphi$ is a predicate applied to a mapping from 
column names to values. In sorting, we assume that there is a permutation on row indices $q$ such
that the tuples obtained by selecting values according to the given sort key are ordered. The 
auxiliary definition $u_{k,l}$ negates the number to reverse the sort order when descending order
is required.

The most complex operation is grouping. We need to group data by the value of the column $f_i$ and
then apply aggregations $\rho_1,\ldots,\rho_m$. To do this, we first obtain a set of groups $g_1,\ldots, g_s$
where each group represents a set of indices of rows belonging to each group. For a given group 
$g_i$ we can then obtain values of column $j$ for rows in the group as $\{v_{j,k}\,|\, k\in g_i\}$.
This is used to calculate the resulting data set -- the field $f_i$ becomes a new column formed
by the group keys (obtained by picking one of the indices from $g_j$ for each group); other fields
are calculated by aggregating data in various ways -- $|g_i|$ gives the number of rows in the group,
$\Sigma$ sums numerical values and $\Pi$ (a slight notation abuse) concatenates string values.

@@

\subparagraph{Reduction rules.} The reduction relation $\leadsto_L$ is parameterized by a function
$L$ that maps class names to class definitions, together with nested classes associated with the
class definition (used during type checking as discussed below). The map is not used in the 
reduction rules for the relational algebra, given in Figure~\ref{fig:foo-relsem} and so it was 
omitted there.

@@

A type provider is an executable component called by the compiler and the editor to 
provide information about types on demand. In our formalization, we follow the 
style of Petricek et al. \cite{fsdata}, but we add laziness as discussed in Section~\ref{sec:foo-foo}.
We model the core operations (dropping columns, grouping and sorting) in Section~\ref{sec:pivot-core}
and refine the model to include filtering Section~\ref{sec:pivot-filter}.
For simplicity we omit paging, which does not affect the shape of data.

@@

A type provider is a function that takes static parameters, such as schema of the input data set, 
and returns a class name $C$ together with a mapping that defines the body of the class and 
definitions of delayed classes $L$ that may be used by the members of the class $C$. 
In our case, the schema $F$ is a mapping from field names to field types:
%
\begin{equation*}
\begin{array}{l}
\ident{pivot}(F) = C, \{ C \mapsto (\kvd{type}~C(x:\ident{Query}) = \ldots,L) \} \quad \textnormal{where}~F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}
\end{array}
\end{equation*}
%
The class $C$ provided by the pivot type provider has a constructor taking \ident{Query}, which 
represents the, possibly already partly transformed, input data set. It generates members that
allow the user to refine the query and access the data. The type provider is defined using several
helper functions discussed in the rest of this section.

@@

If we were using the relational algebra formalized in Section~\ref{sec:foo-rel} to construct 
queries, we can write an invalid program, e.g.~by attempting to select
a column $f$ using $\Pi_{f}$ from a data set that does not contain the column. This is not an issue 
when using the pivot type provider, because the provided types allow the user to construct only 
correct data transformations.

To formalize this, we prove partial soundness of the Foo calculus (Theorem~\ref{thm:foo-sound}),
which characterizes the invalid programs that can be written using the \ident{Query}-typed 
expressions and then prove safety of the pivot type provider (Theorem~\ref{thm:pivot-safe}),
which shows that such errors do not occur when using the provided types.

\subparagraph{Foo calculus.} The Foo calculus consists of the relational algebra and simple object
calculus where objects can be constructed and their members accessed. It permits recursion as a
member can invoke itself on a new object instance. To accommodate this, we formalize soundness 
using progress (Lemma~\ref{thm:foo-progress}) and preservation (Lemma~\ref{thm:foo-pres}).

@@

The example given in Section~\ref{sec:intro} obtained top 8 athletes based on the number
of gold medals from Rio 2016. It used two operations that were omitted in the formalization in
Section~\ref{sec:pivot-core}. We omitted paging to keep the host language simple, but we also
omitted filtering, which lets us write $\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}$.
This operation is worth further discussion. To support it, the type provider needs not only the
schema of the data set, but also sample data set that is used to offer the available values such 
as $\qident{Rio (2016)}$.

In the revised formalization, the \ident{pivot} function which models the type provider takes the schema
$F$ together with sample data $D$ and provides the type with class context:
%
\begin{equation*}
\begin{array}{l}
\ident{pivot}(F, D) = C, L \quad \textnormal{where}\\
\qquad F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}\\
\qquad D=\{ f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} 
\end{array}
\end{equation*}

@@

\begin{figure}[b]
\begin{center}
\includegraphics[scale=0.30,trim=0mm 0mm 0mm 0mm,clip]{images/table.png} % left bottom right top
\end{center}
\caption{Athletes by the number of medals over the entire history of Olympic games}
\label{fig:case-table}
\end{figure}

@@

We used The Gamma script with the pivot type provider to build an interactive web site 
(\url{rio2016.thegamma.net}) that visualizes a number of facts about Olympic medalists 
using the data set discussed in Appendix~\ref{app:olympics-csv} and used throughout this paper.
The web site lets the readers view and modify the source code and we also developed a number of
tools that make working with the source code easier, going beyond the basic auto-completion tooling
to enable dot-driven development as discussed in Section~\ref{sec:analysis-auto}. In this section,
we review our experience and outline some of the additional tools (available at 
\url{github.com/the-gamma}).

@@

The user interface lets the user choose aggregations to be calculate over a group and select 
columns used for sorting. It is generated automatically by looking for specific pattern in the 
chain of member accesses -- we annotate members with annotations denoting whether a member is 
start of a list, list item or an end of a list. The editor then looks for parts of the chain of the
form $\qident{list start}.\qident{list item 1}.\qident{list item 2}.\qident{list end}$ and generates
a component that lets the user remove or add list items. An item cannot be removed if the operation
would break the code (e.g.~when it adds a member that is needed later) and items to be added
are chosen using available members (as in the standard auto-complete). The headers
shown in Figure~\ref{fig:case-opts} are provided as additional annotations attached to 
\qident{list start}.

@@

The Figure~\ref{fig:case-ed} shows the sample expression (discussed above) in the live 
editor\footnote{The live editor can be tested live as part of the documentation for the 
JavaScript package at \url{thegamma.net}}.
Note that the selected part of code is the \qident{by Gold descending} identifier and so the 
preview shows results as computed at that point of the query evaluation. Athletes with largest number
of gold medals appear first, but silver or bronze medals are not yet used as secondary sorting keys.
As the user moves through the code, or writes the code, the live preview is updated accordingly.

@@

\subparagraph{Human-computer interaction.}
We discussed how the pivot type provider simplifies the programming model (Section~\ref{sec:analysis}),
but it would be interesting to explore this aspect empirically through the perspective of HCI. 
The live editor shown in Section~\ref{sec:impl} offers a form of direct manipulation 
\cite{directman,directman2,dynamicq}. Unlike spreadsheets, we construct a transformation rather than
actually transforming data, which makes it more related to systems for query construction
\cite{spreadsheetalgebra,querydirect}. Our approach is somewhat different in that we see code 
as equally important to the direct manipulation interface.

@@

The pivot type provider offers a safe and easy to use layer over an underlying relational algebra 
that we use to model data transformations. As a key technical contribution of this paper, we 
formalize the type provider and prove that queries constructed using the types it provides are 
correct. Achieving this property by other means would require a language with complex
type system features such as typestate and row types.

@@

The data set used as an example in the case study discussed in Section~\ref{sec:impl} as well as
in the examples discussed throughout the paper is a single CSV file listing the entire history of
Olympic medals awarded since 1896. The data set can be found at \url{https://github.com/the-gamma/workyard}
together with scripts used to obtain it. The following is a representative example listing the
first 5 rows:
%
{\small
\begin{equation*}
\begin{array}{l}
\textbf{\sffamily {Games},{Year},{Discipline},{Athlete},{Team},{Gender},{Event},{Medal},{Gold},{Silver},{Bronze}}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Alfred Hajos},\kvd{HUN},\ident{Men},\ident{100m freestyle},\ident{Gold},\num{1},\num{0},\num{0}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Otto Herschmann},\kvd{AUT},\ident{Men},\ident{100m freestyle},\ident{Silver},\num{0},\num{1},\num{0}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Dimitrios Drivas},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors},\ident{Bronze},\num{0},\num{0},\num{1}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Ioannis Malokinis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors},\ident{Gold},\num{1},\num{0},\num{0}\\
\ident{Athens (1896)},\num{1896},\ident{Swimming},\ident{Spiridon Chasapis},\kvd{GRE},\ident{Men},\ident{100m freestyle for sailors},\ident{Silver},\num{0},\num{1},\num{0}\\
\end{array}
\end{equation*} }
